Higher-Order unification for free!

### 2024-04-11 paper/main.tex
Update relaxing progress-llam-fail

\newcommand{\Uo}{\ensuremath{\UnifRel_o}\xspace}
\newcommand{\nUo}{\ensuremath{\nUnifRel_o}\xspace}
\newcommand{\Eo}{\ensuremath{\EqualRel_o}\xspace}
\newcommand{\nEo}{\ensuremath{\nEqualRel_o}\xspace}

@@

\newcommand{\Fo}{\ensuremath{\mathcal{F}_{\!o}\xspace}} % space non va
\newcommand{\Ho}{\ensuremath{\mathcal{H}_o}\xspace}

@@

\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
\newcommand{\foUnifPb}{\ensuremath{\mathbb{P}}\xspace}
\newcommand{\hoUnifPb}{\ensuremath{\mathbb{T}}\xspace}

@@

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}
\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
\newcommand{\notllambda}{\ensuremath{\Diamond\llambda}\xspace}
%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}

@@

\subsection{The problem: logic-program Simulation}
\todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} of length $\mathcal{N}$.
At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
\footnote{If the same rule is used multiple times in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time.}
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho''}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

@@

\subsection{Memory map (\mapStore) and substitution ($\rho$ and $\sigma$)}
\label{sec:grounwork}

@@

\subsection{Definition of \Uo and its properties}\label{sec:founif}

@@

\section{Handling of \notllambda}\label{sec:beta}

@@

\subsection{Relaxing \fullref{def:progBetaFail}}

@@

% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}. 
There exists systems
such as $\lambda$Prolog~\cite{lamProlog} which 
delay the resolution of \notllambda unification problems if the substitution
is not able to put them in \llambda. For example in 
\noteD{Teijus}
%
% make test ONLY=7005 TEX=tex
\printAlll
  {{{(X\appsep a),a},
    {X,\lambda x.Y}}}
  {{}}
  {{}}
  {{}}

In the example above, $\foUnifPb_1$ is in \notllambda and the object language
cannot solve it, and, by \cref{prop:fidelity}, the meta language neither.
However, we can be more permissive, and relax \cref{lemma:keep-fidelity}. This
modification is quite simple to manage: we are introducing a new \notllambda
progress rule, say \progBetaNoLLWait, by which, if \lhs is rigid and \rhs is
flexible, the considered \linkbeta is kept in the store and no progression is
done\footnote{This new rule trivially guarantees the termination of progress}.
\progBetaNoLLWait makes \elpiIn{occur-check-links} partial, since the check is
possible only on links with a variable on the \lhs. This means that we can have
two links \linkbetaM{}{X}{Y\appsep a} and \linkbetaM{}{f\appsep X}{Y\appsep a}
where the occur check does not throw an error. Note however, that the
decompilation of the two links will force the unification of $X$ to $Y\appsep a$
and then the unification of $f\appsep (Y \appsep a)$ to $Y\appsep a$, which
fails by the occur check of \Ue.

A second strategy to deal with problem that are in \notllambda is to make some
approximation. This is the case for example of the unification algorithm of Coq
used in its type class solver~\cite{sozeau08}. The approximation consists in
forcing a choice (among the others) when the unification problem is in
\notllambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
that this is of course an approximation, since $\sigma = \{X \mapsto \lambda x.Y, Y \mapsto
\_\}$ is another valid substitution for the original problem. This approximation
can be easily introduced in our unification procedure, by adding new custom 
\linkbeta progress rules.

% \noteD{Paragraph for decompile in case of \maybebeta}
Decompilation of \linkbeta is possible by extending \elpiIn{commit-link} with
new heuristics.
\todo{cita paper su estensioni a llam via sospensione etc}
\todo{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}

>>>>

\newcommand{\Uo}{\texorpdfstring{\ensuremath{\UnifRel_o}\xspace}{unif\_o}}
\newcommand{\nUo}{\ensuremath{\nUnifRel_o}\xspace}
\newcommand{\Eo}{\ensuremath{\EqualRel_o}\xspace}
\newcommand{\nEo}{\ensuremath{\nEqualRel_o}\xspace}

@@

\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}

@@

\newcommand{\linkStore}{\texorpdfstring{\ensuremath{\mathbb{L}}\xspace}{L}}
\newcommand{\mapStore}{\texorpdfstring{\ensuremath{\mathbb{M}}\xspace}{M}}
\newcommand{\foUnifPb}{\ensuremath{\mathbb{P}}\xspace}
\newcommand{\hoUnifPb}{\ensuremath{\mathbb{T}}\xspace}

@@

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}
\newcommand{\maybeeta}{\texorpdfstring{\ensuremath{\Diamond\eta}\xspace}{maybeeta}}
\newcommand{\maybebeta}{\texorpdfstring{\ensuremath{\Diamond\beta_0}\xspace}{maybebeta}}
\newcommand{\notllambda}{\texorpdfstring{\ensuremath{\Diamond\llambda}\xspace}{maybellam}}
%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}

@@

\subsection{The problem: logic-program simulation}
\todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} of length $\mathcal{N}$.
At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
\footnote{If the same rule is used multiple times in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time.}
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho''}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

@@

\subsection{Memory map (\mapStore) and substitution (\texorpdfstring{$\rho$ and $\sigma$}{rho and sigma})}
\label{sec:grounwork}

@@

\subsection{Definition of \Uo{} and its properties}\label{sec:founif}

@@

\section{Handling of \notllambda}\label{sec:beta}
\noteD[inline]{Dire che OL fallisce tutte le unif not in llambda, e poi modificare 
le progress e la decomp}

@@

\subsection{\texorpdfstring{Relaxing \fullref{def:progBetaFail}}{Relaxing progress-llam-fail}}

@@

% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}. 
There exists systems
such as Teyjus~\cite{Nadathur2001} and $\lambda$Prolog~\cite{lamProlog} which 
delay the resolution of \notllambda unification problems if the substitution
is not able to put them in \llambda.

In this section we want to show how we can adapt the unification of the object
language in the meta language by simply adding (or removing) rules to the
\elpiIn{progress} predicate.
%
% make test ONLY=7005 TEX=tex
\printAlll
  {{{(X\appsep a),a},
    {X,\lambda x.a}}}
  {{}}
  {{}}
  {{}}

In the example above, $\foUnifPb_1$ is in \notllambda. \emph{If} the object
language delays the first unification problem waiting $X$ to be
be instantiated in a future unification, we can relax \cref{def:progBetaFail}. 
Instead of failing because the \lhs of the considered \linkbeta $l$ becomes rigid,
we keep it in \linkStore until the head of its \rhs also become rigid.
In this case, since \lhs and \rhs have rigid heads, they can be unified
just before removing $l$ from \linkStore. We can note that this rule trivially
guarantees \fullref{prop:fidelity}. On the other hand, the occur check becomes 
partial: there exists \linkbeta with a non-flexible \lhs.

A second strategy to deal with problem that are in \notllambda is to make
approximations. This is the case for example of the unification algorithm of Coq
used in its type class solver~\cite{sozeau08}. The approximation consists in
forcing a choice (among the others) when the unification problem is outside
\llambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
that this is of course an approximation, since $\sigma = \{X \mapsto \lambda x.Y, Y \mapsto
\_\}$ is another valid substitution for the original problem. We stress the 
fact that, again, our unification procedure in the meta language can be 
accommodated for this new behavior: given a \linkbeta, if \lhs is not in \llambda,
then \elpiIn{progress} can try to align the rightmost arguments and unify
the resulting heads.

Note that delaying unification outside \llambda can leave \linkbeta during the
decompilation phase. Therefore, new rules to \elpiIn{commit-links} should
be added accordingly.

% \noteD{Paragraph for decompile in case of \maybebeta}
% \todo{cita paper su estensioni a llam via sospensione etc}
\todo[inline]{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}
